Study of Intra- and Inter-Job Interference for Improving Allocation on HPC Systems

### 2015-10-13 submission.tex
minor change

A parallel application usually conforms to a combination of several basic communication patterns\cite{hpdc2015-ornl}. At its different execution phases, the application's communication behavior may follow different basic patterns respectively. Collective communication is often used for data distribution and collection among processes. Nearest Neighbor communication is commonly seen in lots of structured and unstructured mesh physics packages. There are many profiling tools available to capture communication patterns from parallel applications, such as Tuning and Analysis Utilities(TAU)\cite{tau}, mpiP\cite{mpip}. They can help analysis parallel application, providing information like the percentage of different MPI operations of the applications, communication topology, the amount data transferred between processes.

In this work, we provide detailed analysis about the communication pattern of three DOE MiniApps. MiniApps are reduced proxy applications that encapsulate the salient performance of larger full size applications\cite{miniapp}. Since we only focus on the dominant communication pattern of parallel applications, these MiniApps are perfect candidates for analysis the application's performance on different allocations or on new network architectures. We choose three MiniApps, namely, Algebraic Multigrid Solver(AMG), Geometric MultiGrid(MultiGrid) and CrystalRouter. Their representative communication patterns that are commonly seen in the HPC system workload. The communication pattern figures we present here are generated with the IPM\cite{ipm} data from \cite{design forward webpage}.

@@

AMG is the MiniApp of Algebraic Multigrid Solver, which is a parallel algebraic multigrid solver for linear systems arising from problems on unstructured mesh physics packages. It has been derived directly from the BoomerAMG solver that is being developed in the Center for Applied Scientific Computing (CASC) at LLNL\cite{amg}. The dominant communication pattern is this regional communication with decreasing message size for different parts of the multigrid v-cycle.

@@

AMG represents a group of applications whose dominant communication pattern is Nearest Neighbor. Applications with such similar dominant communication pattern are like PARTISN, SNAP\cite{design forward webpage}.

@@

The second MiniApp we have is Crystal Router, which is the extracted communication kernel of the full application Nek5000. Nek5000\cite{crystalrouter} is a spectral element CFD application developed at Argonne National Laboratory. It features spectral element multigrid solvers coupled with a highly scalable, parallel coarse-grid solver that widely used for projects including ocean current modeling, thermal hydraulics of reactor cores, and spatiotemporal chaos. The communication kernel of Nek5000, CrystalRouter demonstrates the "many-to-many" communication pattern through scalable multi-stage communication process. \textcolor{yellow}{The way CrystalRouter works is nodes compute for a while, then synchronize and communicate, continually alternating between these two types of activities.}

@@

\subsection{MultiGrid}
\label{sec:multigrid}
Different from AMG, MultiGrid is geometric multigrid v-cycle from production elliptic solver BoxLib, a software framework for massively parallel block-structured adaptive mesh refinement (AMR) codes. MultiGrid conforms to Many-to-Many communication pattern with decreasing message size and collectives for different parts of the multigrid v-cycle. It is widely used for structured grid physics packages. Figure\ref{fig: mg communication topology} shows the communication topology of MultiGrid and Figure\ref{fig: mg data trans} shows the the amount of data transferred between ranks in MultiGrid.

@@

There are couple of simulation platform to study the behavior of large scale systems. SimGrid from Inria is a versatile simulation tool sets that provide simulation for Grids, Clouds and HPC systems\cite{simgrid}. The Structural Simulation Toolkit (SST) from Sandia National Lab was developed to explore innovations in highly concurrent systems where the ISA, microarchitecture, and memory interact with the programming model and communications system\cite{sst}.

A simulation toolkit named CODES (Co-Design of Multi-layer Exascale Storage Architecture) from Argonne National Lab enables the exploration of simulating different HPC networks with high fidelity\cite{Jason-2011}\cite{mubarak-sc2012}. CODES is built on top of Rensselaer Optimistic Simulation System (ROSS) parallel discrete-event simulator, which is capable of processing billions of events per second on leadership-class supercomputers\cite{ross}. CODES support both torus and dragonfly network with high fidelity flit-level simulation. In this work, we only use torus network to show the impact of node allocation to job with different dominant communication patterns. CODES has this network workload component that capable of conducting trace-driven simulation. It can take real MPI application trace generated by SST DUMPI\cite{sst} to drive CODES network models. We implemented a Job Mapping API for CODES, which can provide flexible job allocation strategies. The new API can help us to explore the impact of different allocation strategies to those three applications in Section \ref{sec:application study}

@@

CODES can provide Torus network model with good scalability and high fidelity\cite{codes-torus}. In this section, we will first study the communication behavior of three applications on torus network with different dimensionality and bandwidth configurations. Then, we study the intra- and inter-job interference with these three applications by running them exclusively and concurrently with different allocation strategies.

@@

Communication variability due to network sharing can cause application performance degradation in two ways, namely intra- and inter-job interference. The \textbf{\emph{intra-job interference}} refers to the network contention between ranks within each application. There are lots of research works\cite{jingjin}\cite{ozan} try to come up with better task mapping algorithms to alleviate the intra-job interference, which is out the scope of our work. In this work, we focus on the analysis of such interference with different allocations.

\textbf{\emph{Inter-job interference}} is introduced by concurrently running jobs that adjacent to each other sharing network resources. Such interference has been identified as one of the major culprits for application's performance variability\cite{abhinav-sc13}\cite{skinner}\cite{rosenthal}. Inter-job interference is a more prominent issue for system adopted non-contiguous allocation policy than partition-based system. Application's communication time can vary from 36\% faster to 69\% slower due to inter-job interference when running on non-contiguous allocation\cite{abhinav-sc13}.

@@

3D unbalanced-cube, shown as green in Figure \ref{fig:cont_sub1}, is a rectangular prism, which is the possible allocation shape on system with asymmetric networks. For example, Blue Waters Cray XE6/XK7 system network is 3D torus with Gemini routers. The network connections in the \emph{y}direction have only half the bandwidth of the cables used in the \emph{x} and \emph{z} directions. In order to take advantage of the faster links in the \emph{x} and \emph{z} directions, job allocation is always start from X-Z plane, which leads to a rectangular prism shaped allocation\cite{RF}.

@@

\emph{OBSERVATION}
Compact allocation should be provided to application with global data transfer, such as those conforms to ``Many-to-Many" communication pattern. Application with dominant communication pattern like Nearest Neighbor won't fully utilize the compact allocation and could get even better performance on a relaxed allocation shape.

@@

\emph{OBSERVATION}
The performance of application conforms to Nearest Neighbor communication pattern keep relatively stable with different allocations. Non-contiguous allocation won't aggravate intra-job interference as long as the unit size can preserve the locality in application's communication topology graph.

@@

\emph{OBSERVATION}
The unit size in non-contiguous allocation policy should be chosen according to application's communication pattern. Inter-job interference is inevitable in non-contiguous based systems, but unit size that big enough to preserve the locality of the application will alleviate such interference and improve job performance.

@@

\section{Related Work}
\label{sec:related_work}
%\textbf{P1: tools for monitoring and profiling}\\
There are some tools available for system monitoring and application profiling on HPC systems. Tools like TAU (Tuning and Analysis) \cite{tau}, SCALASCA\cite{scala} and mpiP\cite{mpip} can capture application's runtime information about application's communication and computation in event traces. However, recognizing the patterns about application's communication from those traces requires lots of extra effort. Some researchers work on the recognition and characterization of parallel application communication patterns. Oak Ridge National Laboratory has this ongoing project about developing a tool set named Oxbow, which can characterize the computation and communication behavior of 12 scientific applications and benchmarks\cite{oxbow}. In their recent work\cite{hpdc2015-ornl}, they demonstrate a new approach to automatically characterizing a parallel application's communication behavior.

%\textbf{P2: communication characteristics}\\
Some research institutes put great effort in characterization of scientific applications. The Department of Energy initiate this design forward project aims at the identification of the computational characteristics of the DOE MiniApps developed at various exascale co-design centers\cite{design forward webpage}. In this project, the communication patterns of several DOE full applications and associated mini-applications are studied to provide a more complete snapshot of the DOE workload. A joint project named CORAL from Oak Ridge, Argonne and Livermore provide a series of benchmarks to represent DOE workloads and technical requirements\cite{coral}. The CORAL project includes scalable science benchmarks, throughput benchmarks, data centric benchmarks, skeleton benchmarks and Micro benchmarks.

%\textbf{P3: Interference between concurrently running jobs}\\
The interference between concurrently running jobs on HPC system have been identified as major culprit for job's performance variability. Abhinav et al. found that concurrently running applications on HPC can cause interference to each other, and cause communication time varied from 36\% faster to 69\% slower. Such interference could come from Operating System noise, shape of the allocated partition and mainly other running jobs that sharing network resources\cite{abhinav-sc13}. Skinner et al. found that  there is a 2-3$\texttimes$ slowdown in MPI$\textunderscore$Allreduce due to network contention from other jobs\cite{skinner}. Rosenthal et al. found there is limited benefit for many applications by increasing network bandwidth. The applications that send mostly small messages or larger messages asynchronously are not bandwidth bounded, hence, benefit only slightly or not at all from increased bandwidth\cite{rosenthal}.

%\textbf{P4: Job allocation(little bit)}\\
There are some research work focus on optimizing job allocation on HPC system to alleviate the interference between concurrently running jobs. Hoefler et al. propose to use performance modeling techniques to analysis factors that impact the performance of parallel scientific applications\cite{hoefler-modeling}. However, as the scale of HPC continue grows, the  interference of concurrently running jobs is getting worse, which is hard to be quantified by performance profiling tools. Bogdan et al provide a set of guidelines how to configure a network with Dragonfly topology for workload with Nearest Neighbor communication pattern\cite{Bogdan-hpdc14}. Dong et al describe IBM Blue Gene/Q's 5D torus interconnect network\cite{Dong-SC11}. They developed simple benchmarks that conforms to four different communication patterns, namely ping-pong, nearest neighbor, broadcast and allreduce, to demonstrate the effectiveness of this highly parallel 5D torus network.

>>>>

A parallel application usually conforms to a combination of several basic communication patterns \cite{hpdc2015-ornl}. At its different execution phases, the application's communication behavior may follow different basic patterns respectively. Collective communication is often used for data distribution and collection among processes. Nearest Neighbor communication is commonly seen in lots of structured and unstructured mesh physics packages. There are many profiling tools available to capture communication patterns from parallel applications, such as Tuning and Analysis Utilities(TAU) \cite{tau}, mpiP \cite{mpip}. They can help analysis parallel application, providing information like the percentage of different MPI operations of the applications, communication topology, the amount data transferred between processes.

In this work, we provide detailed analysis about the communication pattern of three DOE MiniApps. MiniApps are reduced proxy applications that encapsulate the salient performance of larger full size applications \cite{miniapp}. Since we only focus on the dominant communication pattern of parallel applications, these MiniApps are perfect candidates for analysis the application's performance on different allocations or on new network architectures. We choose three MiniApps, namely, Algebraic Multigrid Solver(AMG), Geometric MultiGrid(MultiGrid) and CrystalRouter. Their representative communication patterns that are commonly seen in the HPC system workload. The communication pattern figures we present here are generated with the IPM \cite{ipm} data from \cite{design forward webpage}.

@@

AMG is the MiniApp of Algebraic Multigrid Solver, which is a parallel algebraic multigrid solver for linear systems arising from problems on unstructured mesh physics packages. It has been derived directly from the BoomerAMG solver that is being developed in the Center for Applied Scientific Computing (CASC) at LLNL \cite{amg}. The dominant communication pattern is this regional communication with decreasing message size for different parts of the multigrid v-cycle.

@@

AMG represents a group of applications whose dominant communication pattern is Nearest Neighbor. Applications with such similar dominant communication pattern are like PARTISN, SNAP \cite{design forward webpage}.

@@

The second MiniApp we have is Crystal Router, which is the extracted communication kernel of the full application Nek5000. Nek5000 \cite{crystalrouter} is a spectral element CFD application developed at Argonne National Laboratory. It features spectral element multigrid solvers coupled with a highly scalable, parallel coarse-grid solver that widely used for projects including ocean current modeling, thermal hydraulics of reactor cores, and spatiotemporal chaos. The communication kernel of Nek5000, CrystalRouter demonstrates the "many-to-many" communication pattern through scalable multi-stage communication process. \textcolor{yellow}{The way CrystalRouter works is nodes compute for a while, then synchronize and communicate, continually alternating between these two types of activities.}

@@

\subsection{MultiGrid}
\label{sec:multigrid}
Different from AMG, MultiGrid is geometric multigrid v-cycle from production elliptic solver BoxLib, a software framework for massively parallel block-structured adaptive mesh refinement (AMR) codes. MultiGrid conforms to Many-to-Many communication pattern with decreasing message size and collectives for different parts of the multigrid v-cycle. It is widely used for structured grid physics packages. Figure \ref{fig: mg communication topology} shows the communication topology of MultiGrid and Figure\ref{fig: mg data trans} shows the the amount of data transferred between ranks in MultiGrid.

@@

There are couple of simulation platform to study the behavior of large scale systems. SimGrid from Inria is a versatile simulation tool sets that provide simulation for Grids, Clouds and HPC systems \cite{simgrid}. The Structural Simulation Toolkit (SST) from Sandia National Lab was developed to explore innovations in highly concurrent systems where the ISA, microarchitecture, and memory interact with the programming model and communications system\cite{sst}.

A simulation toolkit named CODES (Co-Design of Multi-layer Exascale Storage Architecture) from Argonne National Lab enables the exploration of simulating different HPC networks with high fidelity \cite{Jason-2011}\cite{mubarak-sc2012}. CODES is built on top of Rensselaer Optimistic Simulation System (ROSS) parallel discrete-event simulator, which is capable of processing billions of events per second on leadership-class supercomputers \cite{ross}. CODES support both torus and dragonfly network with high fidelity flit-level simulation. In this work, we only use torus network to show the impact of node allocation to job with different dominant communication patterns. CODES has this network workload component that capable of conducting trace-driven simulation. It can take real MPI application trace generated by SST DUMPI \cite{sst} to drive CODES network models. We implemented a Job Mapping API for CODES, which can provide flexible job allocation strategies. The new API can help us to explore the impact of different allocation strategies to those three applications in Section \ref{sec:application study}

@@

CODES can provide Torus network model with good scalability and high fidelity \cite{codes-torus}. In this section, we will first study the communication behavior of three applications on torus network with different dimensionality and bandwidth configurations. Then, we study the intra- and inter-job interference with these three applications by running them exclusively and concurrently with different allocation strategies.

@@

Communication variability due to network sharing can cause application performance degradation in two ways, namely intra- and inter-job interference. The \textbf{\emph{intra-job interference}} refers to the network contention between ranks within each application. There are lots of research works \cite{jingjin}\cite{ozan} try to come up with better task mapping algorithms to alleviate the intra-job interference, which is out the scope of our work. In this work, we focus on the analysis of such interference with different allocations.

\textbf{\emph{Inter-job interference}} is introduced by concurrently running jobs that adjacent to each other sharing network resources. Such interference has been identified as one of the major culprits for application's performance variability \cite{abhinav-sc13}\cite{skinner}\cite{rosenthal}. Inter-job interference is a more prominent issue for system adopted non-contiguous allocation policy than partition-based system. Application's communication time can vary from 36\% faster to 69\% slower due to inter-job interference when running on non-contiguous allocation \cite{abhinav-sc13}.

@@

3D unbalanced-cube, shown as green in Figure \ref{fig:cont_sub1}, is a rectangular prism, which is the possible allocation shape on system with asymmetric networks. For example, Blue Waters Cray XE6/XK7 system network is 3D torus with Gemini routers. The network connections in the \emph{y}direction have only half the bandwidth of the cables used in the \emph{x} and \emph{z} directions. In order to take advantage of the faster links in the \emph{x} and \emph{z} directions, job allocation is always start from X-Z plane, which leads to a rectangular prism shaped allocation \cite{RF}.

@@

%\emph{OBSERVATION}
Compact allocation should be provided to application with global data transfer, such as those conforms to ``Many-to-Many" communication pattern. Application with dominant communication pattern like Nearest Neighbor won't fully utilize the compact allocation and could get even better performance on a relaxed allocation shape.

@@

%\emph{OBSERVATION}
The performance of application conforms to Nearest Neighbor communication pattern keep relatively stable with different allocations. Non-contiguous allocation won't aggravate intra-job interference as long as the unit size can preserve the locality in application's communication topology graph.

@@

%\emph{OBSERVATION}
The unit size in non-contiguous allocation policy should be chosen according to application's communication pattern. Inter-job interference is inevitable in non-contiguous based systems, but unit size that big enough to preserve the locality of the application will alleviate such interference and improve job performance.

@@

\section{Related Work}
\label{sec:related_work}
%\textbf{P1: tools for monitoring and profiling}\\
There are some tools available for system monitoring and application profiling on HPC systems. Tools like TAU (Tuning and Analysis) \cite{tau}, SCALASCA \cite{scala} and mpiP \cite{mpip} can capture application's runtime information about application's communication and computation in event traces. However, recognizing the patterns about application's communication from those traces requires lots of extra effort. Some researchers work on the recognition and characterization of parallel application communication patterns. Oak Ridge National Laboratory has this ongoing project about developing a tool set named Oxbow, which can characterize the computation and communication behavior of 12 scientific applications and benchmarks \cite{oxbow}. In their recent work \cite{hpdc2015-ornl}, they demonstrate a new approach to automatically characterizing a parallel application's communication behavior.

%\textbf{P2: communication characteristics}\\
Some research institutes put great effort in characterization of scientific applications. The Department of Energy initiate this design forward project aims at the identification of the computational characteristics of the DOE MiniApps developed at various exascale co-design centers \cite{design forward webpage}. In this project, the communication patterns of several DOE full applications and associated mini-applications are studied to provide a more complete snapshot of the DOE workload. A joint project named CORAL from Oak Ridge, Argonne and Livermore provide a series of benchmarks to represent DOE workloads and technical requirements \cite{coral}. The CORAL project includes scalable science benchmarks, throughput benchmarks, data centric benchmarks, skeleton benchmarks and Micro benchmarks.

%\textbf{P3: Interference between concurrently running jobs}\\
The interference between concurrently running jobs on HPC system have been identified as major culprit for job's performance variability. Abhinav et al. found that concurrently running applications on HPC can cause interference to each other, and cause communication time varied from 36\% faster to 69\% slower. Such interference could come from Operating System noise, shape of the allocated partition and mainly other running jobs that sharing network resources \cite{abhinav-sc13}. Skinner et al. found that  there is a 2-3$\texttimes$ slowdown in MPI$\textunderscore$Allreduce due to network contention from other jobs \cite{skinner}. Rosenthal et al. found there is limited benefit for many applications by increasing network bandwidth. The applications that send mostly small messages or larger messages asynchronously are not bandwidth bounded, hence, benefit only slightly or not at all from increased bandwidth \cite{rosenthal}.

%\textbf{P4: Job allocation(little bit)}\\
There are some research work focus on optimizing job allocation on HPC system to alleviate the interference between concurrently running jobs. Hoefler et al. propose to use performance modeling techniques to analysis factors that impact the performance of parallel scientific applications \cite{hoefler-modeling}. However, as the scale of HPC continue grows, the  interference of concurrently running jobs is getting worse, which is hard to be quantified by performance profiling tools. Bogdan et al provide a set of guidelines how to configure a network with Dragonfly topology for workload with Nearest Neighbor communication pattern \cite{Bogdan-hpdc14}. Dong et al describe IBM Blue Gene/Q's 5D torus interconnect network \cite{Dong-SC11}. They developed simple benchmarks that conforms to four different communication patterns, namely ping-pong, nearest neighbor, broadcast and allreduce, to demonstrate the effectiveness of this highly parallel 5D torus network.
